String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-26 2023-1-19/sections/cefa.tex
algorithm

In this section, we recall the definition of CEFA and Parikh image.

%\subsection{Cost-Enriched Finite Automaton} \label{subsec:cefa}
The definition of CEFA in this paper is lightly different from the definition in \cite{atva2020}. In \cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a 7-tuple $(Q, \Sigma, \delta, q_I, F, R, \theta)$ where
  \begin{itemize}
    \item $Q,\Sigma,q_I,F$ is defined as NFA,
    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta$ is a transition set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\theta: F\rightarrow \varphi$ is a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
  The value of each register $r_i$ in $R$ is written as $\mathcal{V}(r_i)$ and is initiated to 0 at the initial state. A \emph{run} of $\aut$ on string $a_1\cdots a_m$ is a transition sequence $q_I\xrightarrow[\myvec{v_1}]{a_1}q_1\cdots q_{m-1}\xrightarrow[\myvec{v_m}]{a_m}q_m$  and $\mathcal{V}(r_i) = \displaystyle\sum_{j=1}^m \myvec{v_j}[i], i\in [1,n]$ is the value of $r_i$ after the run. $\theta[R/\mathcal{V}(R)](q_m)$ is obtained from $\theta(q_m)$ by replacing each register $r_i$ to its value $\mathcal{V}(r_i)$. The run is \emph{accepting} if $q_m\in F$ and $\theta[R/\mathcal{V}(R)](q_m)$ is satisfiable. $\top$ is the valid formula that is always satisfiable. A string $w$ is accepted by $\aut$ if it has an accepting run of $\aut$. The language of $\aut$ is the set of strings accepted by $\aut$, denoted by $\lan(\aut)$.

\end{definition}
\begin{example} \label{eg:1}
  Fig.\ref{fig:repetition} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm,
          accepting text=$1\leq r_1\leq 100$, accepting/.style=accepting by arrow}
      ]
      \node[state,initial]            (q_0)                      {$q_0$};
      \node[state]                    (q_1) [right=of q_0]       {$q_1$};
      \node[state,accepting]          (q_2) [right=of q_1]       {$q_2$};

      \path[->] (q_0) edge              node      [above]           {$a$/(1)} (q_1)
      (q_1) edge              node      [above]           {$b$/(0)} (q_2)
      (q_2) edge [bend left]  node      [below]           {$\epsilon$/(0)} (q_0);
    \end{tikzpicture}
    \caption{An automaton recognizing the language of $(ab)\{1,100\}$}
    \label{fig:repetition}
  \end{figure}
\end{example}
\begin{example}[The pre-image of length operation]\label{eg:pre_len} The length operation can be captured by a CEFA base on our previous work \cite{atva2020}. Fig.\ref{fig:len} shows a pre-image of linear literal $i=|x|$. It is easy to see that the pre-image accepts string words with length $i$.
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm},
      accepting text=${r_1=1}$, accepting/.style=accepting by arrow
      ]

      \node[state,initial,accepting]            (q_0)       {$q$};

      \path[->] (q_0) edge [loop below] node{$\Sigma$/(1)} ();
    \end{tikzpicture}
    \caption{The pre-image of $i = |x|$}
    \label{fig:len}
  \end{figure}
\end{example}

>>>>

In this section, we introduce the definition of CEFA, Parikh image and pre-iamge.

%\subsection{Cost-Enriched Finite Automaton} \label{subsec:cefa}
The definition of CEFA in this paper is lightly different from the definition in \cite{atva2020}. In \cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]\newline
  A cost-enriched finite automaton $\aut$ is a 7-tuple $(Q, \Sigma, \delta, q_I, F, R, \theta)$ where
  \begin{itemize}
    \item $Q,\Sigma,q_I,F$ is defined as NFA,
    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta$ is a transition set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\theta: F\rightarrow \varphi$ is a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
  The value of each register $r_i$ in $R$ is written as $\mathcal{V}(r_i)$ and is initiated to 0 at the initial state. A \emph{run} of $\aut$ on string $a_1\cdots a_m$ is a transition sequence $q_I\xrightarrow[\myvec{v_1}]{a_1}q_1\cdots q_{m-1}\xrightarrow[\myvec{v_m}]{a_m}q_m$  and $\mathcal{V}(r_i) = \displaystyle\sum_{j=1}^m \myvec{v_j}[i], i\in [1,n]$ is the value of $r_i$ after the run. $\theta[R/\mathcal{V}(R)](q_m)$ is obtained from $\theta(q_m)$ by replacing each register $r_i$ to its value $\mathcal{V}(r_i)$. The run is \emph{accepting} if $q_m\in F$ and $\theta[R/\mathcal{V}(R)](q_m)$ is satisfiable. $\top$ is the valid formula that is always satisfiable. A string $w$ is accepted by $\aut$ if it has an accepting run of $\aut$. The language of $\aut$ is the set of strings accepted by $\aut$, denoted by $\lan(\aut)$.

\end{definition}
\begin{example} \label{eg:1}
  Fig.\ref{fig:repetition} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm,
          accepting text=$\scriptstyle 1\leq r_1\leq 100$, accepting/.style=accepting by arrow}
      ]
      \node[state,initial]            (q_0)                      {$q_0$};
      \node[state]                    (q_1) [right=of q_0]       {$q_1$};
      \node[state,accepting]          (q_2) [right=of q_1]       {$q_2$};

      \path[->] (q_0) edge              node      [above]           {$\scriptstyle a:(1)$} (q_1)
      (q_1) edge              node      [above]           {$\scriptstyle b:(0)$} (q_2)
      (q_2) edge [bend left]  node      [below]           {$\scriptstyle\epsilon: (0)$} (q_0);
    \end{tikzpicture}
    \caption{An automaton recognizing the language of $(ab)\{1,100\}$}
    \label{fig:repetition}
  \end{figure}
\end{example}
